---
title: "Data Extraction and Clean up"
author: "Benzon Carlitos Salazar (salazarbc24@uww.edu)"
date: "`r Sys.Date()`"
output: pdf_document
---

The data cleanup involved reading raw CSV files from the ORCHID dataset, transforming them into a structured format, and 
saving the cleaned data as Parquet files. The key steps included:

### 1. Reading Raw Data:

- CSV files for various event categories (e.g., ABG, CBC, Chemistry, Culture, Fluid Balance, Hemodynamics, etc.) were read into R.

### 2. Data Cleaning and Transformation:

- Unnecessary columns (`RowID`, unnamed columns) were dropped.
- Data was grouped by event type (e.g., `abg_name`, `cbc_name`, `chem_name`) and converted to a wide format using `pivot_wider()`.
- Numeric values were coerced into appropriate data types (`double`).
- An `opo_group` identifier was extracted from `PatientID` to categorize patients.
- `tidyr::fill()` was used to propagate missing values up and down within each group.
- Data was grouped by `PatientID`, `time_event`, and other relevant identifiers, then deduplicated using `slice_head()`.

### 3. Data Output & Cleanup:

- The cleaned datasets were saved in Parquet format for efficient storage and processing.
- Raw CSV files were removed after ensuring successful cleanup.
- Memory was cleared using `rm(list = ls(all.names = TRUE))` and `gc()` to optimize performance.

This process ensured that data was well-structured, deduplicated, and efficiently stored while maintaining data integrity.

Below is an example of what the whole clean up process looked like.

```{r setup}
library(magrittr)

orchid_folder <- here::here("data-raw/physionet.org/files/orchid/2.0.0")
```

```{r data_cleaning__abg_events, eval=FALSE, include=TRUE}
abg_events <- 
  readr::read_csv(here::here(orchid_folder, "ABGEvents.csv"))

abg_events_proc <-
  abg_events %>%
  dplyr::select(-c("...1", "RowID")) %>%
  dplyr::group_by(abg_name) %>%
  dplyr::mutate(row_id = dplyr::row_number()) %>%
  tidyr::pivot_wider(names_from = abg_name, values_from = value) %>%
  dplyr::ungroup() %>%
  dplyr::select(-row_id) %>%
  dplyr::mutate(opo_group = as.factor(substring(PatientID, 0, 4)),
                abg_ventilator_mode = as.character(abg_ventilator_mode),
                PH = as.double(PH),
                PCO2 = as.double(PCO2),
                PO2 = as.double(PO2),
                HCO3 = as.double(HCO3),
                BE = as.double(BE),
                O2SAT = as.double(O2SAT),
                FIO2 = as.double(FIO2),
                Rate = as.double(Rate),
                TV = as.double(TV),
                PEEP = as.double(PEEP),
                PIP = as.double(PIP))

abg_events_final <-
  abg_events_proc %>%
  dplyr::group_by(PatientID, time_event, abg_ventilator_mode, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

abg_events_final %>% nanoparquet::write_parquet(here::here("data", "abg_events.parquet"))
rm(abg_events, abg_events_proc, abg_events_final)
```

```{r data_cleaning__calc_deaths, eval=FALSE, include=FALSE}
cal_deaths <- 
  readr::read_csv(here::here(orchid_folder, "calc_deaths.csv"))

cal_deaths %>% nanoparquet::write_parquet(here::here("data", "calc_deaths.parquet"))
rm(cal_deaths)
```

```{r data_cleaning__cbc_events, eval=FALSE, include=FALSE}
cbc_events <- 
  readr::read_csv(here::here(orchid_folder, "CBCEvents.csv"))

cbc_events_proc <-
  cbc_events %>%
  dplyr::select(-c("...1", "RowID")) %>%
  dplyr::group_by(cbc_name) %>%
  dplyr::mutate(row_id = dplyr::row_number()) %>%
  tidyr::pivot_wider(names_from = cbc_name, values_from = value) %>%
  dplyr::ungroup() %>%
  dplyr::select(-row_id) %>%
  dplyr::mutate(opo_group = as.factor(substring(PatientID, 0, 4)),
                RBC = as.double(RBC),
                WBC = as.double(WBC),
                Hgb = as.double(Hgb),
                Hct = as.double(Hct),
                Ptl = as.double(Ptl),
                Segs = as.double(Segs),
                Lymp = as.double(Lymp),
                Mono = as.double(Mono),
                Eos = as.double(Eos))

cbc_events_final <-
  cbc_events_proc %>%
  dplyr::group_by(PatientID, time_event, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

cbc_events_final %>% nanoparquet::write_parquet(here::here("data", "cbc_events.parquet"))
rm(cbc_events, cbc_events_proc, cbc_events_final)
```

```{r data_cleaning__chem_events, eval=FALSE, include=FALSE}
chem_events <- 
  readr::read_csv(here::here(orchid_folder, "ChemistryEvents.csv"))

chem_events_proc <-
  chem_events %>%
  dplyr::select(-c("...1", "RowID")) %>%
  dplyr::group_by(chem_name) %>%
  dplyr::mutate(row_id = dplyr::row_number()) %>%
  tidyr::pivot_wider(names_from = chem_name, values_from = value) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(opo_group = as.factor(substring(PatientID, 0, 4))) %>%
  dplyr::select(-row_id)

chem_events_final <-
  chem_events_proc %>%
  dplyr::group_by(PatientID, time_event, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

chem_events_final %>% nanoparquet::write_parquet(here::here("data", "chem_events.parquet"))
rm(chem_events, chem_events_proc, chem_events_final)
```

```{r data_cleaning__culture_events, eval=FALSE, include=FALSE}
culture_events <-
  readr::read_csv(here::here(orchid_folder, "CultureEvents.csv"))

culture_events_proc <-
  culture_events %>%
  dplyr::select(-c("...1", "RowID")) %>%
  dplyr::group_by(culture_source) %>%
  dplyr::mutate(rowid = dplyr::row_number()) %>%
  tidyr::pivot_wider(names_from = culture_source, values_from = result) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(opo_group = as.factor(substring(PatientID, 0, 4))) %>%
  dplyr::select(-rowid)

culture_events_final <-
  culture_events_proc %>%
  dplyr::group_by(PatientID, time_event, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

culture_events_final %>% nanoparquet::write_parquet(here::here("data", "culture_events.parquet"))
rm(culture_events, culture_events_proc, culture_events_final)
```

```{r data_cleaning__fb_events, eval=FALSE, include=FALSE}
fluid_balance_events <-
  readr::read_csv(here::here(orchid_folder, "FluidBalanceEvents.csv"))

fluid_balance_events_proc <-
  fluid_balance_events %>%
  tidyr::pivot_wider(names_from = fluid_name, values_from = amount) %>%
  dplyr::mutate(opo_group = as.factor(substring(PatientID, 0, 4))) %>%
  dplyr::select(-c("...1", "RowID"))

fluid_balance_events_final <-
  fluid_balance_events_proc %>%
  dplyr::group_by(PatientID, time_event_start, time_event_end, opo_group, fluid_type) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

fluid_balance_events_final %>% nanoparquet::write_parquet(here::here("data", "fluid_balance_events.parquet"))
rm(fluid_balance_events, fluid_balance_events_proc, fluid_balance_events_final)
```

```{r data_cleaning__hemo_events, eval=FALSE, include=FALSE}
hemo_events <-
  readr::read_csv(here::here(orchid_folder, "HemoEvents.csv"))

hemo_events_proc <-
  hemo_events %>%
  dplyr::mutate(measurement_name_type = paste0(measurement_type, measurement_name),
                opo_group = as.factor(substring(PatientID, 0, 4))) %>%
  tidyr::pivot_wider(names_from = measurement_name_type, values_from = value) %>%
  dplyr::select(-c("...1", "RowID", "measurement_name", "measurement_type"))

hemo_events_final <-
  hemo_events_proc %>%
  dplyr::group_by(PatientID, time_event_start, time_event_end, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

hemo_events_final %>% nanoparquet::write_parquet(here::here("data", "hemo_events.parquet"))
rm(hemo_events, hemo_events_proc, hemo_events_final)
```

```{r data_cleaning__referrals, eval=FALSE, include=FALSE}
referrals <-
  readr::read_csv(here::here(orchid_folder, "referrals.csv"))

referrals %>% nanoparquet::write_parquet(here::here("data", "referrals.parquet"))
rm(referrals)
```

```{r data_cleaning__serology_events, eval=FALSE, include=FALSE}
serology_events <-
  readr::read_csv(here::here(orchid_folder, "SerologyEvents.csv"))

serology_events_proc <-
  serology_events %>%
  dplyr::mutate(
    serology_name = ifelse(is.na(serology_name_Other), serology_name, paste(serology_name, serology_name_Other, sep = "_")),
    opo_group = as.factor(substring(PatientID, 0, 4)),
    result = paste(result, "SEROLOGY", sep = "_")
  ) %>%
  tidyr::pivot_wider(names_from = result, values_from = serology_name) %>%
  dplyr::select(-c("...1", "RowID", "serology_name_Other"))

serology_events_final <-
  serology_events_proc %>%
  dplyr::group_by(PatientID, time_event, opo_group) %>%
  tidyr::fill(dplyr::everything(), .direction = "updown") %>%
  dplyr::slice_head() %>%
  dplyr::ungroup()

serology_events_final %>% nanoparquet::write_parquet(here::here("data", "serology_events.parquet"))
rm(serology_events, serology_events_proc, serology_events_final)
```

```{r clean_up}
# List of files to remove
files_to_remove <- c(
  "ABGEvents.csv",
  "calc_deaths.csv",
  "CBCEvents.csv",
  "ChemistryEvents.csv",
  "CultureEvents.csv",
  "FluidBalanceEvents.csv",
  "HemoEvents.csv",
  "referrals.csv",
  "SerologyEvents.csv"
)

# Remove files only if they exist
purrr::walk(files_to_remove, ~ {
  file_path <- here::here(orchid_folder, .x)
  if (file.exists(file_path)) {
    file.remove(file_path)
  }
})

rm(list = ls(all.names = TRUE)) # clear all objects including hidden objects
invisible(gc()) # free up memory
```

## Purpose of Data Cleanup

The raw ORCHID dataset contains a wide variety of clinical event data stored across multiple CSV files. These files are 
structured in a long format with multiple redundant columns, inconsistent naming conventions, and varying data types. 
The primary objective of the cleanup was to create tidy, analysis-ready data by:

- Converting event-level long-form data into patient-level wide format.
- Ensuring consistency across numeric, categorical, and time-based variables.
- Removing duplicate rows, irrelevant columns, and noise.
- Storing processed outputs in a space-efficient, analysis-friendly format (`.parquet`).

This enables downstream applications such as predictive modeling, visualization, and exploratory data analysis to 
proceed without further data wrangling.

## Overview of Files Processed

| Raw File Name             | Description                          | Output Filename                  |
|--------------------------|--------------------------------------|----------------------------------|
| `ABGEvents.csv`          | Arterial blood gas measurements      | `abg_events.parquet`             |
| `CBCEvents.csv`          | Complete blood count lab events      | `cbc_events.parquet`             |
| `ChemistryEvents.csv`    | Chemistry lab test results           | `chem_events.parquet`            |
| `CultureEvents.csv`      | Microbial culture results            | `culture_events.parquet`         |
| `FluidBalanceEvents.csv` | Fluid input/output documentation     | `fluid_balance_events.parquet`   |
| `HemoEvents.csv`         | Hemodynamic monitoring values        | `hemo_events.parquet`            |
| `SerologyEvents.csv`     | Serological test outcomes            | `serology_events.parquet`        |
| `calc_deaths.csv`        | Calculated patient mortality data    | `calc_deaths.parquet`            |
| `referrals.csv`          | Referral information (specialty etc.)| `referrals.parquet`              |

## Notes on Data Quality and Consistency

- **Column Removal**: Columns such as `RowID` and unnamed index columns (e.g., `...1`) were dropped to reduce noise.
- **Wide Format Transformation**: Measurements grouped by name (e.g., `abg_name`, `cbc_name`) were converted into wide 
format using `pivot_wider()` for one-row-per-timepoint structure.
- **Type Coercion**: Key clinical variables were cast into `double` to ensure numeric operations are consistent.
- **Group-level Missingness Imputation**: `tidyr::fill()` was used within each group (e.g., per `PatientID`) to forward- and backward-fill missing values where reasonable.
- **Deduplication Strategy**: Within each patient-time-event grouping, only the first row was retained using `slice_head()`, assuming earlier rows are most relevant.
- **Group Identifier**: An `opo_group` variable was extracted using the first four characters of `PatientID` to allow stratified subgroup analyses.

## Folder Structure and Output Format

- Raw data is located in: `data-raw/physionet.org/files/orchid/2.0.0/`
- Cleaned files are saved to: `data/`
- All output files are written in **Parquet** format using the `nanoparquet::write_parquet()` function for fast I/O and minimal disk usage.
