---
title: 'Objective 1: EDA'
author: "Benzon Carlitos Salazar (salazarbc24@uww.edu)"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
library(magrittr)

referrals_df <- nanoparquet::read_parquet(here::here("data", "referrals.parquet"))
```

## Objective 1: Analyze Disparities in Organ Procurement Outcomes

#### Goal

To assess whether organ procurement rates differ across demographic and geographic groups—particularly by race/ethnicity, 
socioeconomic proxies, and OPO region—and quantify the magnitude and significance of these disparities.

* **Outcome Variable**
  * `organ_procured` (binary): 1 if organ was successfully procured, 0 otherwise.
* **Primary Exposure Variables**
  * **Age**
  * **Gender**
  * **Race**
  * **Organ Procurement Organizations (OPOs)**
* **Covariates (for adjustment)**
  * UNOS defined cause of death
  * UNOS defined mechanism of death
  * UNOS defined circumstances of death
  * Time of referral from hospital

### Organ Procurement Outcomes by Demographic, Geographic, and Clinical Characteristics

```{r table1, echo=FALSE}
# obj1_abt %>% dplyr::group_by(Cause_of_Death_UNOS) %>% dplyr::summarise(counts = dplyr::n(), .groups = "keep") %>% dplyr::arrange(desc(counts))

age_cats <-
  c("Less than 18 y.o.",
    "18-19 y.o.",
    "20-29 y.o.",
    "30-39 y.o.",
    "40-49 y.o.",
    "50-59 y.o.",
    "60-69 y.o.",
    "70+ y.o.")

obj1_abt <-
  referrals_df %>%
  dplyr::filter(!is.na(Age) & !is.na(Gender)) %>%
  dplyr::mutate(
    age_category = factor(
      dplyr::case_when(
        Age < 18 ~ age_cats[1],
        dplyr::between(Age, 18, 19) ~ age_cats[2],
        dplyr::between(Age, 20, 29) ~ age_cats[3],
        dplyr::between(Age, 30, 39) ~ age_cats[4],
        dplyr::between(Age, 40, 49) ~ age_cats[5],
        dplyr::between(Age, 50, 59) ~ age_cats[6],
        dplyr::between(Age, 60, 69) ~ age_cats[7],
        Age >= 70 ~ age_cats[8],
        TRUE ~ "Error"
      ),
      levels = age_cats
    ),
    Cause_of_Death_UNOS = ifelse(is.na(Cause_of_Death_UNOS), "Unknown Cause of Death", Cause_of_Death_UNOS),
    Cause_of_Death_UNOS = ifelse(Cause_of_Death_UNOS %in% c("Anoxia", "Unknown Cause of Death", "Other", "CVA/Stroke", "Head Trauma"), Cause_of_Death_UNOS, "Other Cause of Death"),
    Mechanism_of_Death = ifelse(is.na(Mechanism_of_Death), "Unknown Mechanism of Death", Mechanism_of_Death),
    Mechanism_of_Death = ifelse(Mechanism_of_Death %in% c("Unknown Mechanism of Death", "Cardiovascular", "Natural Causes", "ICH/Stroke", "None of the Above", "Blunt Injury"), Mechanism_of_Death, "Other Mechanism of Death"),
    Circumstances_of_Death = dplyr::case_when(
      is.na(Circumstances_of_Death) ~ "Unknown Circumstances of Death",
      Circumstances_of_Death %in% c("Natural Causes", "Death from Natural Causes") ~ "Death from Natural Causes",
      Circumstances_of_Death %in% c("MVA", "Motor Vehicle Accident") ~ "Motor Vehicle Accident",
      TRUE ~ "Other Circumstances of Death"
    )
  ) %>%
  dplyr::select(
    procured,
    Age,
    age_category,
    Gender,
    Race,
    OPO,
    Cause_of_Death_UNOS,
    Mechanism_of_Death,
    Circumstances_of_Death,
    time_referred
  )

obj1_abt_row <-
  obj1_abt %>%
  dplyr::select(-time_referred) %>%
  gtsummary::tbl_summary(
    by = procured,
    percent = "row",
    statistic = list(
      Age ~ "{mean} ({p25}, {p75})",
      gtsummary::all_categorical() ~ "{n}/{N} ({p}%)"
    ),
    sort = list(-age_category ~ "frequency"),
    label = list(
      Age ~ "Mean Patient Age (IQR)",
      age_category ~ "Age Category - no. (%)",
      Gender ~ "Gender - no. (%)",
      Race ~ "Race - no. (%)",
      OPO ~ "Organ Procurement Organization - no. (%)",
      Cause_of_Death_UNOS ~ "UNOS defined cause of death - no. (%)",
      Mechanism_of_Death ~ "UNOS defined mechanism of death - no. (%)",
      Circumstances_of_Death ~ "UNOS defined circumstances of death - no. (%)"
    )
  ) %>%
  gtsummary::modify_header(label = "**Variable**",
                           stat_1 = "**Not Procured**\n(N = {n})",
                           stat_2 = "**Procured**\n(N = {n})") %>%
  gtsummary::bold_labels() %>%
  gtsummary::modify_footnote(update = dplyr::everything() ~ NA)

obj1_abt_col <-
  obj1_abt %>%
  dplyr::select(-c(time_referred, procured)) %>%
  gtsummary::tbl_summary(
    statistic = list(
      Age ~ "{mean} ({p25}, {p75})",
      gtsummary::all_categorical() ~ "{n}/{N} ({p}%)"
    ),
    sort = list(-age_category ~ "frequency"),
    label = list(
      Age ~ "Mean Patient Age (IQR)",
      age_category ~ "Age Category - no. (%)",
      Gender ~ "Gender - no. (%)",
      Race ~ "Race - no. (%)",
      OPO ~ "Organ Procurement Organization - no. (%)",
      Cause_of_Death_UNOS ~ "UNOS defined cause of death - no. (%)",
      Mechanism_of_Death ~ "UNOS defined mechanism of death - no. (%)",
      Circumstances_of_Death ~ "UNOS defined circumstances of death - no. (%)"
    )
  ) %>%
  gtsummary::modify_header(gtsummary::all_stat_cols() ~ "**Overall**\n(N = {N})") %>%
  gtsummary::bold_labels() %>%
  gtsummary::modify_footnote(update = dplyr::everything() ~ NA)

obj1_abt_full <-
  gtsummary::tbl_merge(list(obj1_abt_col, obj1_abt_row)) %>%
  gtsummary::modify_caption("Organ Procurement vs. Outcomes") %>%
  gtsummary::modify_spanning_header(dplyr::everything() ~ NA)

obj1_abt_full %>%
  gtsummary::as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header", "scale_down"),
    stripe_color = "gray!15"
  ) %>%
  kableExtra::column_spec(1, width = "12.5em")
```

Table \@ref(tab:table1) summarizes the baseline characteristics of the study cohort (N = 132,968), stratified by 
whether an organ was successfully procured. Among all decedents, 9,502 (7.1%) had at least one organ procured, while 
123,466 (92.9%) did not.

**Age** was strongly associated with organ procurement. The median age among those whose organs were not procured was 59 
years (IQR: 50–72), compared to 40 years (IQR: 28–54) among those who underwent procurement. Stratified age group analysis 
revealed that younger patients had higher procurement rates; for instance, decedents aged 20–29 had a procurement rate of 
25%, compared to only 16% among those under 18. Procurement likelihood declined progressively with increasing age, 
suggesting a potential age-related disparity that may reflect clinical ineligibility, comorbidity burden, or differences 
in referral or donor evaluation practices.

**Gender** differences in procurement were modest. Males comprised 59% of the cohort and had a procurement rate of 7.5%, 
compared to 6.7% among females. Although the absolute difference was small, this slight imbalance may stem from anatomical 
compatibility, clinical profiles, or unmeasured systemic factors. However, without adjustment for clinical variables, it 
is unclear whether this difference reflects a true disparity or confounding.

**Race and ethnicity** exhibited more pronounced disparities. White/Caucasian decedents accounted for 60% of the cohort, 
with a procurement rate of 7.3%. In contrast, Black/African American decedents (19% of the cohort) had a lower procurement 
rate of 6.1%, while Hispanic decedents (16%) had a notably higher rate of 8.5%. The lower procurement among Black patients 
may reflect structural inequities in referral, consent, or evaluation processes. Conversely, the relatively high procurement 
among Hispanic decedents may be influenced by geographic or demographic differences in donor identification and suitability. 
These trends highlight the importance of adjusting for clinical and regional context in disparity analyses.

**Organ Procurement Organization (OPO)** performance varied substantially across regions. Procurement rates ranged from 
5.3% in OPO3 to 10.2% in OPO5. While differences in case mix may explain some variation, the consistently higher yield in 
OPO5 suggests operational or procedural differences in donor management and recovery efficiency. Conversely, lower rates 
in OPOs 2, 3, and 6 may reflect logistical challenges, under-referral, or systemic inefficiencies. These inter-regional 
disparities underscore the need for standardized performance evaluation and the potential for best-practice dissemination 
from higher-performing OPOs.

**UNOS-defined cause of death** was reported for all decedents. The most common causes were **anoxia** (30%) and 
**unknown cause of death** (22%). **Head trauma** had one of the highest procurement rates at **24%**, followed by 
**CVA/stroke** at **13%**, suggesting these conditions often preserve organ viability. In contrast, **other** and 
**unknown** causes of death had substantially lower procurement rates, **1.0%** and **0%**, respectively. The complete 
lack of procurement among decedents with an unknown cause of death highlights the potential impact of documentation gaps 
on organ recovery. Collectively, these patterns underscore the importance of both clinical suitability and accurate case 
classification in maximizing donation opportunities.

**UNOS-defined mechanism of death** was available for all decedents. The most frequent mechanisms were **unknown** (26%), 
**cardiovascular** (21%), and **natural causes** (19%). Procurement rates were highest among decedents with 
**other mechanisms of death** (27%) and **blunt injury** (22%), both of which are consistent with external trauma and 
greater clinical suitability for organ donation. In contrast, **natural causes** and **cardiovascular deaths** were 
associated with low procurement rates, **1.2%** and **6.1%**, respectively—likely reflecting comorbidities or rapid 
physiological deterioration. Notably, **no organs were procured** from the 34,500 patients (26%) with an 
**unknown mechanism of death**, underscoring how incomplete documentation may contribute to missed donation opportunities.

**UNOS-defined circumstances of death** further highlight these patterns. **Death from natural causes** was the most 
common circumstance (48%) but had a relatively low procurement rate of **5.8%**. In contrast, **motor vehicle accidents**, 
while accounting for only 4.1% of the cohort, were associated with a substantially higher procurement rate of **25%**, 
indicating the clinical viability of donors who experience sudden traumatic deaths. **Other circumstances of death**, a 
heterogeneous category comprising 22% of cases, had a procurement rate of **15%**, suggesting mixed donor suitability. 
As with other variables, **no procurement occurred among the 34,444 patients (26%) with unknown circumstances of death**, 
emphasizing the critical need for complete and accurate documentation to improve identification of potential donors.

```{r gg-procured-by-obj1, fig.cap = "Proportion of organ procurement by Race, Gender, Age Category, and OPO. Bars show within-group percentages; procurement is more common among younger, male, Hispanic, and OPO5 patients.", echo=FALSE}
p_gender <-
  ggplot2::ggplot(obj1_abt, ggplot2::aes(x = Gender, fill = procured)) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::scale_fill_manual(
    values = c("FALSE" = "#D3D3D3", "TRUE" = "#4682B4"),
    name = "Procured",
    labels = c("No", "Yes")
  ) +
  ggplot2::labs(title = "By Gender", y = "Percent", x = NULL) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal()

p_race <-
  ggplot2::ggplot(obj1_abt, ggplot2::aes(x = Race, fill = procured)) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::scale_fill_manual(
    values = c("FALSE" = "#D3D3D3", "TRUE" = "#4682B4"),
    name = "Procured",
    labels = c("No", "Yes")
  ) +
  ggplot2::labs(title = "By Race", y = "Percent", x = NULL) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal()

p_age <-
  ggplot2::ggplot(obj1_abt, ggplot2::aes(x = age_category, fill = procured)) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::scale_fill_manual(
    values = c("FALSE" = "#D3D3D3", "TRUE" = "#4682B4"),
    name = "Procured",
    labels = c("No", "Yes")
  ) +
  ggplot2::labs(title = "By Age Category", y = "Percent", x = NULL) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal()

p_opo <-
  ggplot2::ggplot(obj1_abt, ggplot2::aes(x = OPO, fill = procured)) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::scale_fill_manual(
    values = c("FALSE" = "#D3D3D3", "TRUE" = "#4682B4"),
    name = "Procured",
    labels = c("No", "Yes")
  ) +
  ggplot2::labs(title = "By OPO", y = "Percent", x = NULL) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal()

library(patchwork)

procured_by_obj1 <-
  (p_race | p_gender) /
  (p_age | p_opo) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

procured_by_obj1
```

Figure \@ref(fig:gg-procured-by-obj1) summarizes disparities in organ procurement rates across key demographic and 
regional groups. Notable patterns include higher procurement rates among younger individuals, males, Hispanic patients, 
and those referred within OPO5. These descriptive trends motivated subsequent hypothesis testing and regression analysis.

### Hypothesis testing

We hypothesize **whether organ procurement is associated with specific demographic, geographic, or clinical characteristics.**

More specifically, we are tesing the following:

- H_{0} (null): There is no difference in organ procurement rates across the levels of the variable.
- H_{1} (alternative): There is a difference in organ procurement rates across the levels of the variable.

While my initial hypothesis testing plan included both demographic and clinical variables, I am excluding `Cause_of_Death_UNOS`, 
`Mechanism_of_Death`, and `Circumstances_of_Death` due to evidence of documentation bias. These variables are 
disproportionately complete among patients whose organs were procured, but often missing or marked as "unknown" for 
those who were not. Because this documentation typically occurs during or after donor evaluation, the completeness of 
these fields may be contingent on the procurement process itself. As a result, statistical associations involving these 
variables may reflect process-related bias rather than true underlying clinical differences.

I also excluded `age_category`, as it is a derived variable from `Age`, which was already included in the analysis 
using a Welch two-sample t-test. Including both would introduce redundancy and increase the risk of multiple comparisons 
without providing additional insight.

My final hypothesis testing focused on variables that are consistently recorded and not conditional on procurement 
activity: `Race`, `Gender`, `OPO`, and continuous `Age`.

```{r hypothesis_testing_2, echo=FALSE}
categorical_vars <- c("Race", "Gender", "OPO")

chi_sq_results <-
  categorical_vars %>%
  stats::setNames(nm = .) %>%
  purrr::map_df(~ {
    tbl <- table(obj1_abt$procured, obj1_abt[[.x]])
    broom::tidy(chisq.test(tbl)) %>%
      dplyr::mutate(variable = .x)
  })

age_test <- 
  broom::tidy(t.test(Age ~ procured, data = obj1_abt)) %>%
  dplyr::mutate(variable = "Age")

final_results <- 
  dplyr::bind_rows(chi_sq_results, age_test) %>%
  dplyr::select(variable, method, statistic, p.value)

final_results
```

All variables tested showed statistically significant differences in procurement rates:

* **Race** was significantly associated with procurement status (*x^{2}* = 126, *p* < 0.001), suggesting that procurement 
rates vary across racial and ethnic groups.
* **Gender** also showed a significant difference (*x^{2}* = 29.7, *p* < 0.001), although the magnitude of the association 
was smaller.
* **Organ Procurement Organization (OPO)** region was strongly associated with procurement likelihood (*x^{2}* = 516, 
*p* < 0.001), indicating considerable variation in outcomes across geographic regions.
* **Age**, treated as a continuous variable, was significantly different between procured and non-procured groups 
(Welch *t* = 103, *p* < 0.001), with younger patients more likely to have organs procured.

These results support the presence of disparities in procurement outcomes related to race, gender, region, and age, 
motivating further multivariable modeling to quantify these differences while adjusting for potential confounding.

### Logistic Regression

```{r log_regression_obj1}
log_reg_model <- stats::glm(procured ~ Race + Gender + OPO, data = obj1_abt, family = binomial)
```

```{r or_obj1, echo=FALSE}
log_reg_results <- broom::tidy(log_reg_model, exponentiate = TRUE, conf.int = TRUE)
log_reg_results %>% knitr::kable()
```

After controlling for all covariates in the model:

* **Race**: Compared to Black patients (reference group), Hispanic patients had **higher odds** of organ procurement 
(OR is approx. 1.35), while patients categorized as Other/Unknown had **lower odds** (OR is approx. 0.87). White/Caucasian 
patients also had significantly higher odds (OR is approx. 1.17), suggesting potential racial disparities in procurement 
outcomes even after adjustment.
* **Gender**: Male patients had **higher odds** of procurement compared to females (OR is approx. 1.12), consistent with 
patterns observed in the unadjusted analysis.
* **OPO Region**: Substantial geographic variation was observed. Compared to OPO1 (reference), patients in OPO5 had the 
**highest odds** of procurement (OR is approx. 1.27), while those in OPO2, OPO3, and OPO6 had **significantly lower odds** 
(ORs ranging from 0.62 to 0.67). These differences may reflect regional variation in practices, infrastructure, or consent 
processes.

All associations were statistically significant (p < 0.05), and none of the confidence intervals included the null value 
(OR = 1.0). This indicates meaningful adjusted differences in organ procurement across race, gender, and OPO.

Specifically, Hispanic patients, White patients, male patients, and those in OPO5 had significantly higher odds of organ 
procurement compared to their respective reference groups. In contrast, patients in OPOs 2, 3, 4, and 6 had significantly 
lower odds.

```{r ors-from-log-reg-model, fig.cap = "Odds Ratios from Logistic Regression Model", echo=FALSE}
log_reg_results %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term = gsub("Race", "Race: ", term),
    term = gsub("Gender", "Gender: ", term),
    term = gsub("OPO", "OPO: ", term),
    term = factor(term, levels = rev(term))
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = term, y = estimate)) +
  ggplot2::geom_point(size = 3, color = "steelblue") +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = conf.low, ymax = conf.high),
    width = 0.1,
    color = "steelblue"
  ) +
  ggplot2::geom_hline(yintercept = 1,
                      linetype = "dashed",
                      color = "red") +
  ggplot2::coord_flip() +
  ggplot2::labs(x = NULL,
                y = "Odds Ratio (95% CI)") +
  ggplot2::theme_minimal(base_size = 12)
```

Figure \@ref(fig:ors-from-log-reg-model) displays the adjusted odds ratios (ORs) and 95% confidence intervals from 
the logistic regression model evaluating associations between demographic/geographic characteristics and the likelihood 
of organ procurement.

```{r clean_up, include=FALSE}
rm(list = ls(all.names = TRUE)) # clear all objects including hidden objects
invisible(gc()) # free up memory
```
