---
title: "Clinical Predictors of Procurement Success"
author: "Benzon Carlitos Salazar (salazarbc24@uww.edu)"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
library(magrittr)

referrals_df <- nanoparquet::read_parquet(here::here("data", "referrals.parquet"))

abg_events_df <- 
  nanoparquet::read_parquet(here::here("data", "abg_events.parquet")) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::arrange(desc(time_event), .by_group = TRUE) %>%
  dplyr::slice_head() %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(opo_group, time_event)) %>%
  dplyr::inner_join(referrals_df %>% dplyr::select(PatientID), by = "PatientID") %>%
  dplyr::distinct()
cbc_events_df <- 
  nanoparquet::read_parquet(here::here("data", "cbc_events.parquet")) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::arrange(desc(time_event), .by_group = TRUE) %>%
  dplyr::slice_head() %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(opo_group, time_event)) %>%
  dplyr::inner_join(referrals_df %>% dplyr::select(PatientID), by = "PatientID") %>%
  dplyr::distinct()
chemisty_events_df <-
  nanoparquet::read_parquet(here::here("data", "chem_events.parquet")) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::arrange(desc(time_event), .by_group = TRUE) %>%
  dplyr::slice_head() %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(opo_group, time_event)) %>%
  dplyr::inner_join(referrals_df %>% dplyr::select(PatientID), by = "PatientID") %>%
  dplyr::distinct() %>%
  dplyr::rename("Sodium" = "NA")
hemo_events_df <-
  nanoparquet::read_parquet(here::here("data", "hemo_events.parquet")) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::arrange(desc(time_event_start), .by_group = TRUE) %>%
  dplyr::slice_head() %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(opo_group, time_event_start, time_event_end)) %>%
  dplyr::inner_join(referrals_df %>% dplyr::select(PatientID), by = "PatientID") %>%
  dplyr::distinct()
fluid_balance_events_df <-
  nanoparquet::read_parquet(here::here("data", "fluid_balance_events.parquet")) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::arrange(desc(time_event_start), .by_group = TRUE) %>%
  dplyr::slice_head() %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(opo_group, time_event_start, time_event_end)) %>%
  dplyr::inner_join(referrals_df %>% dplyr::select(PatientID), by = "PatientID") %>%
  dplyr::distinct()

all_labs_df <-
  referrals_df %>%
  dplyr::select(PatientID, procured, OPO) %>%
  dplyr::mutate(procured = factor(ifelse(procured == TRUE, 1, 0))) %>%
  dplyr::inner_join(abg_events_df, by = "PatientID") %>%
  dplyr::inner_join(cbc_events_df, by = "PatientID") %>%
  dplyr::inner_join(chemisty_events_df, by = "PatientID") %>%
  dplyr::inner_join(fluid_balance_events_df, by = "PatientID") %>%
  dplyr::inner_join(hemo_events_df, by = "PatientID") %>%
  dplyr::select(-value_modifier)
```

## Objective 2: Evaluating the Influence of Clinical Variables on Procurement Success

#### Goal

To evaluate whether specific medical and laboratory parameters, such as blood gas values, chemistry panels, complete 
blood counts, and hemodynamic indicators, are associated with successful organ procurement, and to quantify the strength 
and direction of these associations using univariate logistic regression models.

For this analysis, the following datasets are used:

- Arterial blood gas measurements
- Complete blood count lab events
- Chemistry lab test results 
- Fluid input/output documentation 
- Hemodynamic monitoring values

For each measurement, we take the patient's **latest** value. And then all latest labs information are joined together to 
create a one `all_labs_info` dataframe.

**Outcome**

- `procured` (binary): 1 = organ procured, 0 = not procured.

Since not all patients had every laboratory test completed, the dataset size decreased substantially after joining all lab 
datasets: from an initial 133,101 patients to 10,843 patients with at least one lab result recorded. Importantly, some 
patients may have only partially completed a given panel, for instance, a patient might have a White Blood Cell (WBC) count 
but lack values for Red Blood Cell (RBC) count or Hemoglobin (Hgb). Such patients are still retained in the 
`all_labs_info` dataframe, provided they have any available lab values.

This approach allows us to maximize sample size while preserving partial clinical information, ensuring that we do not 
exclude potentially informative cases due to isolated missingness.

Since there are still a lot of missing values, however, further processing is required to run the logistic regression model.
I will further filter out variables with more than 50% missingness:

```{r missing_work, echo=FALSE}
missing_summary <- 
  all_labs_df %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~mean(is.na(.)))) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "missing_pct")

keep_vars <- 
  missing_summary %>%
  dplyr::filter(missing_pct < 0.5) %>%
  dplyr::pull(variable)

final_df <- 
  all_labs_df %>%
  dplyr::select(all_of(keep_vars)) %>%
  dplyr::select(-c(PatientID, fluid_type)) %>%
  dplyr::mutate(across(where(is.character), as.factor)) %>%
  dplyr::select(dplyr::where(~ dplyr::n_distinct(.) > 1)) %>%
  dplyr::filter(complete.cases(.))
```

After this, we are now at 10,843 observations with 44 usable variables. Additionally, due to the massive imbalance of the 
`fluid_type`, I also had to remove that column, which leaves us with 43 variables.

**Statistical Modeling Approach**:

To identify clinical laboratory variables associated with successful organ procurement, we use a **manual backward stepwise logistic regression** 
approach. We begin with a full model that includes all available lab-based predictors from each patient's most recent 
arterial blood gas (ABG), complete blood count (CBC), chemistry, hemodynamic, and fluid balance data.

At each step, we remove the variable that contributes the least to the model, typically the one with the highest p-value 
or smallest effect size. After each removal, we refit the model and compare it to the previous version using 
**likelihood ratio tests (ANOVA)** or changes in the **Akaike Information Criterion (AIC)**. We continue this process 
until all remaining variables are either statistically significant or clinically relevant.

This approach yields a more parsimonious model, reduces overfitting, and helps isolate the strongest clinical predictors 
of procurement success.

### Multivariable Logistic Regression Identifying Clinical Predictors of Procurement

```{r fit-log-reg-backwards, include=FALSE, message=FALSE, warning=FALSE}
full_model <- stats::glm(procured ~ ., data = final_df, family = binomial)
step_model <- step(full_model, direction = "backward", trace = TRUE)
step_model_summary <- broom::tidy(step_model, exponentiate = TRUE, conf.int = TRUE)
```

```{r log-reg-backwards, echo=FALSE}
knitr::kable(step_model_summary, digits = 3,
             caption = "Final Logistic Regression Model: Clinical Predictors of Procurement Success (Odds Ratios)")
```

To evaluate the association between clinical variables and the likelihood of organ procurement, I performed a multivariable 
logistic regression with backward stepwise elimination. The final model retained significant predictors after adjusting 
for potential collinearity and missingness.

Table \@ref(tab:log-reg-backwards) shows that several variables were independently associated with increased odds of procurement. 
These included higher **arterial pH** (*OR* = 13.1, 95% CI: 1.75–98.0, *p* = 0.012), **white blood cell count** (*OR* = 1.05 
per unit increase, 95% CI: 1.02–1.09, *p* < 0.001), **hemoglobin** (*OR* = 1.45, 95% CI: 1.04–2.05, *p* = 0.033), **sodium** 
(*OR* = 1.04, 95% CI: 1.03–1.07, *p* < 0.001), **albumin** (*OR* = 1.44, 95% CI: 1.11–1.88, *p* = 0.007), and **magnesium** 
(*OR* = 1.58, 95% CI: 1.09–2.34, *p* = 0.020).

In contrast, higher values of **hematocrit** (*OR* = 0.85, 95% CI: 0.76–0.94, *p* = 0.004), **platelet count** (*OR* = 0.99 
per unit increase, 95% CI: 0.993–0.997, *p* < 0.001), **potassium** (*OR* = 0.74, 95% CI: 0.58–0.96, *p* = 0.021), and 
**serum creatinine** (*OR* = 0.90, 95% CI: 0.84–0.96, *p* = 0.002) were negatively associated with procurement.

Ventilator mode was also significant: compared to the reference group, **pressure-controlled ventilation (PC)** was associated 
with increased odds (*OR* = 1.49, 95% CI: 1.01–2.24, *p* = 0.049), whereas **CPAP mode** was associated with decreased odds 
(*OR* = 0.12, 95% CI: 0.012–0.82, *p* = 0.042). Procurement likelihood varied significantly by OPO, with **OPO2** exhibiting 
a markedly higher odds of procurement (*OR* = 2.88, 95% CI: 1.91–4.44, *p* < 0.001) relative to the reference group.

These findings suggest that a combination of metabolic, hematologic, and respiratory parameters, as well as operational 
region, play important roles in procurement success.

```{r clean_up, include=FALSE}
rm(list = ls(all.names = TRUE)) # clear all objects including hidden objects
invisible(gc()) # free up memory
```
