---
title: "Investigating Disparities, Clinical Variables, and Predictive Modeling in Organ Procurement"
author: "Benzon Carlitos Salazar"
date: '`r strftime(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    number_sections: true
    fig_caption: true
  pdf_document:
    df_print: kable
  html_document:
    code_folding: show
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    theme: cosmo
---

\newpage

```{r setup, include=FALSE}
options(knitr.kable.NA = "")
options(knitr.duplicate.label = "allow")
options(gt.html_tag_check = FALSE)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, include=FALSE}
files <- list.files(pattern = "\\d+.*\\Rmd", recursive = TRUE)
chunks <- paste0('```{r child ="', files, '"}\n```\n')
cat(chunks, sep = "\n")
```

# Introduction

Organ transplantation is a critical component of modern healthcare, offering life-saving interventions for patients with 
end-stage organ failure. However, the process of organ procurement, where we have donor referral to successful retrieval, 
is influenced by a complex interplay of medical, logistical, and systemic factors. Despite ongoing efforts to improve 
equity and efficiency in organ donation, disparities persist across racial, socioeconomic, and geographic lines, raising 
concerns about fairness and resource allocation.

My full code for this project can be found here: https://github.com/carrliitos/orchid_investigation

```{r child ="00_Project-Overview.Rmd"}
```

# Data Extraction and Clean up

**All code chunks will not be evaluated since clean up was already performed prior to knitting this document.**

```{r child ="01_Data-Extraction.Rmd"}
```

# Analysis and Results

```{r child ="02_Objective1_EDA.Rmd"}
```

```{r child ="03_Objective2.Rmd"}
```

# Conclusions

This analysis of over 130,000 decedents from the ORCHID dataset reveals clear and statistically significant disparities 
in organ procurement across racial, gender, age, and geographic lines. Younger age, male sex, and residence within certain 
Organ Procurement Organizations (notably OPO5) were all associated with higher odds of organ procurement. After adjusting 
for covariates, both Hispanic and White decedents had significantly greater odds of organ procurement compared to Black 
decedents, indicating persistent racial inequities even when controlling for observable clinical and demographic factors.

Geographic disparities were also pronounced, with some OPO regions demonstrating procurement rates nearly double those of 
others, suggesting variation in operational efficiency, consent practices, or systemic access. Importantly, fields such as 
cause, mechanism, and circumstances of death were often incomplete for non-procured cases, raising the possibility of 
documentation bias and reinforcing the need for standardization and completeness in clinical reporting.

These findings underscore the urgent need for policy-level interventions aimed at standardizing referral practices, 
improving donor identification workflows, and addressing structural barriers to equitable organ donation. I believe that 
future work should focus on disentangling institutional, clinical, and societal drivers of these disparities and 
evaluating targeted interventions across under-performing regions and demographic groups.

# Discussion

The results of my project have important implications for the donor-eligible population represented in the ORCHID dataset. 
The observed disparities in organ procurement rates, especially by race, age, gender, and geographic region (proxied by OPO), 
highlight structural and systemic inequities that may limit access to life-saving transplant opportunities. For example, 
the significantly lower procurement rates among Black decedents, even after adjusting for demographic and geographic factors, 
suggest the influence of unmeasured social, institutional, or consent-related barriers. Likewise, substantial variation 
in procurement performance across OPO regions indicates a lack of consistency in donor identification and recovery practices,
with some regions demonstrating markedly lower efficiency or success rates.

If I were to repeat my study, several improvements could enhance its robustness and validity. First, more comprehensive 
adjustment for clinical variables, such as comorbidities, hemodynamic stability, or documented contraindications to 
procurement, would allow for a more nuanced understanding of whether disparities are clinically justified or reflect 
systemic bias. Second, future analyses would benefit from linkage to hospital- and OPO-level operational data, which 
could show some light on institutional differences in referral timing, staffing, consent protocols, and donor management 
practices. Additionally, integrating qualitative or sociocultural data, such as language preference or next-of-kin 
refusal rates, may help contextualize demographic disparities.

Several potential biases may affect these findings. The most prominent is **documentation bias**—key variables such as 
cause, mechanism, and circumstances of death were often missing or marked as "unknown" in non-procured cases. This likely 
reflects a post-referral or post-evaluation data entry process, where full clinical documentation is completed only after 
a patient is deemed a viable donor (i.e., the donor has passed away). Consequently, analyses that include these fields may 
suffer from **informative missingness**, where the absence of data is correlated with the outcome itself. In addition, 
there may be **selection bias** if referral practices differ systematically across hospitals, regions, or patient 
populations, potentially underrepresenting certain groups in the donor evaluation process.

In summary, while the findings point to real and actionable disparities in organ procurement, they must be interpreted 
in the context of the limitations presented above. Future study on the ORCHID dataset should aim to improve data completeness, 
incorporate richer clinical context, and examine upstream processes, such as referral timing and evaluation criteria, to 
more precisely target opportunities for reform.

\newpage
# Appendix
\listoftables
\listoffigures
\newpage
